Compilation and Backend-Independent Vectorization for Multi-Party Computation

### 2022-04-30 overleaf/backend.tex
Edits to section 6

We require that all inputs are marked as either shared or plaintext, however, we do not require any other qualifier annotations in the program. Then we infer if intermediate variables and expressions are shared through taint analysis with "taintedness" referring to the shared attribute.  Specifically, the compiler follows the following rules, which are standard in positive-negative qualifier systems (here \texttt{shared} is the positive qualifier and \texttt{plain} is the negative one):
\begin{enumerate}
\item Loop counters and array upper bounds are always \texttt{plain}.
\item If any variable on the right-hand side of an assignment is shared, then the assigned variable is \texttt{shared}.
%\item If all variables on the right-hand side of an assignment are plaintext, then the assigned variable is plaintext \ana{Will be covered by the last rule.}
\item Any variables that cannot be determined as shared via the above rules are \texttt{plain}.
\end{enumerate}

@@

MOTION supports FOR loops and SIMD operations, so translation from MPC source to MOTION C++ code is relatively straightforward.
% Variable naming
\paragraph{Variable declarations:}
Our compiler performs taint analysis on MPC Source based on the type system in~\secref{sec:syntax}. The analysis is described in the following section;
for the most part, it is a standard taint analysis that propagates the declared \texttt{shared} qualifiers throughout the routine.
Taint analysis assigns \texttt{shared} or \texttt{plain} qualifiers to all local variables, which is important for code generation.
Our generated C++ uses the following variable-naming scheme: shared variables are named the same as in the MPC Source with the {\sf !} replaced with an underscore (e.g. {\sf sum!2} would be translated to {\sf sum\_2}). Plaintext variables follow the same naming convention as shared variables but are prefixed with {\sf \_MPC\_PLAINTEXT\_}.  The shared representation of constants are named {\sf \_MPC\_CONSTANT\_} followed by the literal constant (e.g. the shared constant 0 would be named {\sf \_MPC\_CONSTANT\_0}).

% Function preamble
\paragraph{Code generation:}
The generated MOTION code begins with the declaration of all variables used in the function, including loop counters.  If a variable is a vectorized array, it is initialized to a correctly-sized array of empty MOTION shares.  Additionally, each plaintext variable and parameter has a shared counterpart declared.  Next, all constant values which are used as part of shared expressions are initialized as a shared input from party 0.  Finally, plaintext parameters are converted used as shared inputs from party 0 to initialize their shared counterparts. (Importantly, as we mention, they are converted only if they are needed as an argument to a shared operation; one can think of it as an upcast from \texttt{plain} to \texttt{share}.)

% Non-vectorized assignments and returns
Once the function preamble is complete, the MPC Source is translated into C++ one statement at a time. The linear structure of MPC Source enables translation. If there is no vectorization present in a statement, translation to C++ is straightforward: outside of MUX statements and array updates, non-vectorized assignments, expressions, and returns directly translate into their C++ equivalents.  Non-vectorized MUX statements are converted to MOTION's MUX member function on the condition variable.  Array updates are translated into two C++ assignments: one to update the value in the original array and one to assign the new array as shown in Listing~\ref{tab:motion_translation_array_updates}.

@@

% Phi nodes and FOR loops
Pseudo PHI nodes are broken into two components: the ``FALSE'' branch which assigns the initial value of the PHI node and the ``TRUE'' branch which assigns the PHI node's back-edge.  The assignment of the false branch occurs right before the PHI node's enclosing loop.  Inside of the PHI node's enclosing loop, a C++ \texttt{if} statement is inserted to only assign the true branch of the PHI node after the first iteration. MPC FOR loops are converted to C++ FOR loops which iterate the loop counter over the specified range.  Listing~\ref{tab:motion_translation_for_loop} illustrates. The loop counter is initialized before any PHI nodes in the loop as they could rely on it; there is no other special handling of FOR loops. %\ana{Add the reference to the corresponding figure somewhere in this paragraph.}

@@

% SIMD representation
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional {\sf std::vector}s in row-major order. %\ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} \ben{MOTION does only support 1-dim SIMD arrays, but since we use helper functions to create and access SIMD arrays this restriction doesn't apply to our representation. There is no reason why these arrays need to be 1-dimensional other than that it slightly simplifies the code generation and avoids some nasty templates on the utility functions.} \ana{Ok! Leave as is.}
Whenever a SIMD value is used in an expression, the utility function {\sf vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression.
%\ana{Is it correct to say that the operators are overloaded? If yes, use that term?} \ben{I don't think that's the right term since SIMD and scalar MOTION values use the same C++ class, so the operator is the same.} \ana{Ok, leave as is.}
Therefore, the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{True. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now but leave the comments too because it needs work.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the {\sf lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  This function evaluates the expression for each permutation of indices along the dimensions and returns the resulting array in row-major order.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the {\sf lift()} function which occurs when we are raising an array.  In this case, instead of concatenating the array for each index, we extend the array along all dimensions being raised which are not present in the array already.  For example, when raising an array with dimensions $N \times M$ to an array with dimensions $N \times M \times D$, the input array will simply be extended along the $D$ dimension: {\sf A'[n,m,d] = A[n,m]} for every $d$.  If the input array is already correctly sized it will be returned as-is.
%This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).
%\ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.} \ben{Rewrote the above paragraph to be less confusing/implementation-specific, please review.} \ana{Better, removing comment.}

@@

% drop_dim() implementation
\paragraph{Dropping dimensions:}
Dropping dimensions use the {\sf drop\_dim()} and {\sf drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.

% Shared/plaintext conversions
\ben{Ishaq or Vassilis: please check over this paragraph... I don't want to make some mistake about MOTION's capabilities.}
Currently, our compiler only supports the \texttt{Bmr} and \texttt{BooleanGMW} protocols as MOTION does not implement all operations for other protocols.  MOTION does not support publicly-known constants for these protocols, so all conversions from plaintext values to shares are performed by providing the plaintext value as a shared input from party 0.  Due to this limitation, our translation to MOTION code attempts to minimize the number of conversions from a plaintext value. This is accomplished by creating a shared copy of each plaintext variable and updating that copy in lock-step with the plaintext variable.  Since variables are often initialized to a common constant value (e.g. 0), this approach decreases the number of input gates by only creating a shared input for each initialization constant.  Loop counters must still be converted to a shared value on each iteration that they are used, however we only generate this conversion when necessary, i.e., when the counter flows to a shared computation. This is to prevent unnecessary increase in the number of input gates.

>>>>

We require that all inputs are marked as either shared or plaintext, however, we do not require any other qualifier annotations in the program. Then we infer if intermediate variables and expressions are shared through taint analysis with "taintedness" referring to the shared attribute.  Specifically, the compiler follows the following rules, which are standard in positive-negative qualifier systems (here \texttt{shared} is the positive qualifier and \texttt{plain} is the negative one):
\begin{enumerate}
\item Loop counters and bounds are always \texttt{plain}.
\item If any variable on the right-hand side of an assignment is shared, then the assigned variable is \texttt{shared}.
%\item If all variables on the right-hand side of an assignment are plaintext, then the assigned variable is plaintext \ana{Will be covered by the last rule.}
\item Any variables that cannot be determined as shared via the above rules are \texttt{plain}.
\end{enumerate}

@@

MOTION supports FOR loops and SIMD operations, so translation from MPC source to MOTION C++ code is relatively straightforward.
\paragraph{Variable declarations:}
% Variable naming
Our generated C++ uses the following variable-naming scheme: shared variables are named the same as in the MPC Source with the {\sf !} replaced with an underscore (e.g. {\sf sum!2} would be translated to {\sf sum\_2}). Plaintext variables follow the same naming convention as shared variables but are prefixed with {\sf \_MPC\_PLAINTEXT\_}.  The shared representation of constants are named {\sf \_MPC\_CONSTANT\_} followed by the literal constant (e.g. the shared constant 0 would be named {\sf \_MPC\_CONSTANT\_0}).

% Function preamble
The generated MOTION code begins with the declaration of all variables used in the function, including loop counters.  If a variable is a vectorized array, it is initialized to a correctly-sized array of empty MOTION shares.  Additionally, each plaintext variable and parameter has a shared counterpart declared.  Next, all constant values which are used as part of shared expressions are initialized as a shared input from party 0.  Finally, plaintext parameters are converted used as shared inputs from party 0 to initialize their shared counterparts. (Importantly, as we mention, they are converted only if they are needed as an argument to a shared operation; one can think of it as an upcast from \texttt{plain} to \texttt{share}.) \ben{The note in parentheses is not true: we currently assume that all plaintext variables other than loop counters will be used in a shared expression.  We initialize a shared version of them in the preamble (using the \texttt{\_MPC\_CONSTANT} variables or the plaintext parameter, depending on how the plaintext variable is initialized) and update that shared variable whenever the plaintext variable is updated.  This is done to minimize the number of input gates (e.g. if a plaintext variable is upcast to shared in a loop, that would insert N input gates).  With protocols that allow publicly known shared constants, this would not be necessary.  This is described in the final subsection of this section.}

\paragraph{Code generation:}
% Non-vectorized assignments and returns
Once the function preamble is complete, the MPC Source is translated into C++ one statement at a time. The linear structure of MPC Source enables this approach to translation. If there is no vectorization present in a statement, translation to C++ is straightforward: outside of MUX statements and array updates, non-vectorized assignments, expressions, and returns directly translate into their C++ equivalents.  Non-vectorized MUX statements are converted to MOTION's MUX member function on the condition variable.  Array updates are translated into two C++ assignments: one to update the value in the original array and one to assign the new array as shown in Listing~\ref{tab:motion_translation_array_updates}.

@@

% Phi nodes and FOR loops
MPC FOR loops are converted to C++ FOR loops which iterate the loop counter over the specified range.  Pseudo PHI nodes are broken into two components: the ``FALSE'' branch which assigns the initial value of the PHI node and the ``TRUE'' branch which assigns the PHI node's back-edge.  The assignment of the false branch occurs right before the PHI node's enclosing loop.  As these assignments may rely on the loop counter, the loop counter is initialized before these statements.  Inside of the PHI node's enclosing loop, a C++ \texttt{if} statement is inserted to only assign the true branch of the PHI node after the first iteration.  Listing~\ref{tab:motion_translation_for_loop} illustrates this translation.%\ana{Add the reference to the corresponding figure somewhere in this paragraph.}

@@

% SIMD representation
\paragraph{Vectorization and SIMD operations:}
Vectorization is handled with utility functions to manage accessing and updating slices of arrays.  All SIMD values are stored in non-vectorized form as 1-dimensional {\sf std::vector}s in row-major order. %\ana{Does MOTION support just 1-dim SIMD arrays? If yes, mention it here.} \ben{MOTION does only support 1-dim SIMD arrays, but since we use helper functions to create and access SIMD arrays this restriction doesn't apply to our representation. There is no reason why these arrays need to be 1-dimensional other than that it slightly simplifies the code generation and avoids some nasty templates on the utility functions.} \ana{Ok! Leave as is.}
Whenever a SIMD value is used in an expression, the utility function {\sf vectorized\_access()} takes the multi-dimentional representation of a SIMD value, along with the size of each dimension and the requested slice's indices, and converts that slice to a MOTION SIMD value. Because MOTION supports SIMD operations using the same C++ operators as non-SIMD operations, we do not need to perform any other transformations to the expression.
%\ana{Is it correct to say that the operators are overloaded? If yes, use that term?} \ben{I don't think that's the right term since SIMD and scalar MOTION values use the same C++ class, so the operator is the same.} \ana{Ok, leave as is.}
Therefore, once vectorized accesses are inserted the translation of an expression containing SIMD values is identical to that of expressions without SIMD values.

@@

\ben{The compiler and C++ internally uses ``lift'' as the terminology for raise\_dim(), but the paper does not :(...} \ana{True. To make matters worse, most of the explanation on this comes in the next section. But I'll leave as is for now but leave the comments too because it needs work.} \ben{Attempted to explain the naming, please review.}
\paragraph{Reshaping and raising dimensions:}
Raising the dimensions of a scalar or array uses the {\sf lift()} utility function which takes a lambda for the raised expression and the dimensions of the output.  This function is also used for the scalar expansion of values which have been lifted out of FOR loops as described in \ref{sec:expansion}.  This function evaluates the expression for each permutation of indices along the dimensions and returns the resulting array in row-major order.  The lambda accepts an array of integers representing the index along each of the dimensions being raised, and the translation of the expression which is being raised replaces each of the dimension index variables with the relevant subscript of this array.  There is also a special case of the {\sf lift()} function which occurs when we are raising an array.  In this case, instead of concatenating the array for each index, we extend the array along all dimensions being raised which are not present in the array already.  For example, when raising an array with dimensions $N \times M$ to an array with dimensions $N \times M \times D$, the input array will simply be extended along the $D$ dimension: {\sf A'[n,m,d] = A[n,m]} for every $d$.  If the input array is already correctly sized it will be returned as-is.
%This case only occurs when the dimensionality of the array is equal to the first dimensions of the array (any dimension which is being lifted to which is not in the array is after every dimension in the array) \ben{I don't know how to phrase the previous sentence more concisely}.  In this case, the array is simply extended along the new dimensions (if the array is already correctly sized it is returned as is).
%\ana{Add the ref. to the figure. Also, try to rewrite. I understand what the paragraph is saying but it is too technical and does not convey the higher-level goal of these. Try first to state what the goal is, and then talk about the details such as the lambda funcs.} \ben{Rewrote the above paragraph to be less confusing/implementation-specific, please review.} \ana{Better, removing comment.}

@@

% drop_dim() implementation
Dropping dimensions use the {\sf drop\_dim()} and {\sf drop\_dim\_monoreturn()} utility functions.  They function identically but the latter returns a scalar for the case when the final dimension of an array is dropped.  These functions take the non-vectorized representation of an array, along with the dimensions of that array, and return the array with the final dimension dropped.

% Shared/plaintext conversions
\paragraph{Upcasting from plaintext to shared:}
Currently, our compiler only supports the \texttt{Bmr} and \texttt{BooleanGMW} protocols as MOTION does not implement all operations for other protocols.  MOTION does not support publicly-known constants for these protocols, so all conversions from plaintext values to shares are performed by providing the plaintext value as a shared input from party 0.  Due to this limitation, our translation to MOTION code attempts to minimize the number of conversions from a plaintext value. This is accomplished by creating a shared copy of each plaintext variable and updating that copy in lock-step with the plaintext variable.  Since variables are often initialized to a common constant value (e.g. 0), this approach decreases the number of input gates by only creating a shared input for each initialization constant.  Loop counters must still be converted to a shared value on each iteration that they are used, however we only generate this conversion when necessary, i.e., when the counter flows to a shared computation. This is to prevent unnecessary increase in the number of input gates when loop counters are only used as plaintext.
\ben{Ishaq or Vassilis: please check over this paragraph... I don't want to make some mistake about MOTION's capabilities.}
